Racial and Ethnic Differences in ETI: Outcomes

### 2024-07-08 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti_outcomes.qmd
data cleaned and tables done

```{r setup}
#| include: false
# Libraries
library(tidyverse)
library(lmerTest)
library(splines)
library(ggeffects)
library(performance)
library(segmented)
# Paths
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data import and minor cleaning}
# Run code for base dataset (takes forever)
# source("/home/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_outcomes_dataset.R")
# Import
load("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/outcomes_dataset.RData")
# Additional quick cleaning steps
# Remove impossible lung function values
encounter$gli_fev1_ppred_rn[encounter$gli_fev1_ppred_rn < 20 |
  encounter$gli_fev1_ppred_rn > 150] <- NA
encounter$gli_fvc_ppred_rn[encounter$gli_fvc_ppred_rn < 20 |
  encounter$gli_fvc_ppred_rn > 150] <- NA
encounter$gli_fev1fvc_ppred_rn[encounter$gli_fev1fvc_ppred_rn < 20 |
  encounter$gli_fev1fvc_ppred_rn > 150] <- NA
# Filter
encounter <- encounter %>%
  filter(
    !is.na(Days),
    Days >= -365.25 * 3 & Days <= 365.25 * 2,
    hospitalized == "No",
    pregnant != 1
  )
```

@@

1. Lung function percent predicted values < 20% or > 150% were excluded.
2. Limited to 3 years before and 2 years after ETI start.
  - Based on `Modulator_trikafta_first_date` variable provided by CFFPR.
3. Hospitalizations and home IV episodes excluded.
4. Years of pregnancy excluded (we only have annualized pregancy data, so don't know exact dates for this).

```{r functions}
# Plots
loess_plot <- function(data = encounter, outcome, facet = F, smooth = "bs",
                       df = 10) {
  p <- ggplot(data, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(
      shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F,
      alpha = 0.1
    ) +
    theme_classic()
  if (smooth == "bs") {
    p <- p + geom_smooth(method = "lm", formula = y ~ splines::bs(x,
      degree = 3, df = df
    ))
  }
  if (smooth == "loess") {
    p <- p + geom_smooth(method = "loess")
  }
  if (smooth == "gam") {
    p <- p + geom_smooth(method = "gam")
  }
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
```

@@

```{r results='asis'}
#| warning: false
#| cache: true
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o, smooth = "loess"))
  print(loess_plot(outcome = o, facet = T, smooth = "loess"))
}))
```

# B-spline smoothing (cubic, df=10)

```{r results='asis'}
#| warning: false
#| cache: true
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o, smooth = "bs"))
  print(loess_plot(outcome = o, facet = T, smooth = "bs"))
}))
```

# B-spline smoothing (cubic, df=3)

```{r results='asis'}
#| warning: false
#| cache: true
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o, smooth = "bs", df = 3))
  print(loess_plot(outcome = o, facet = T, smooth = "bs", df = 3))
}))
```

# GAM smoothing

```{r results='asis'}
#| warning: false
#| cache: true
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o, smooth = "gam"))
  print(loess_plot(outcome = o, facet = T, smooth = "gam"))
}))
```

@@

>>>>

```{r setup}
#| include: false
# Libraries
library(tidyverse)
library(lmerTest)
library(splines)
library(ggeffects)
library(gtsummary)
# Paths
home_dir <- switch(Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF",
  "Linux" = "/home/timvigers/OneDrive/Vigers/CF"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data import}
# Run code for base dataset (takes forever)
# source("/home/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_outcomes_dataset.R")
# Import
load("./Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/outcomes_dataset.RData")
```

@@

1. Lung function percent predicted values < 20% or > 150% were excluded.
  - `r sum(encounter$gli_fev1_ppred_rn < 20,na.rm = T)` ppFEV1 values < 20% and `r sum(encounter$gli_fev1_ppred_rn > 150,na.rm = T)` over 150%.
2. Limited to 3 years before and 2 years after ETI start.
  - Based on `Modulator_trikafta_first_date` variable provided by CFFPR.
3. Hospitalizations and home IV episodes excluded.
  - Based on the care episodes dataset, there were `r length(unique(encounter$eDWID[encounter$hospitalized=="Yes"]))` unique participants with at least one care episode and a total of `r sum(encounter$hospitalized=="Yes")` encounters that were part of a care episode.
  - After filtering out these episodes, there were still 914 encounters with `encounterlocation` variable indicating either hospitalization or home IV, so these were also excluded.
4. Years of pregnancy excluded (we only have annualized pregnancy data, so don't know exact dates for this).
  - `r length(unique(encounter$eDWID[encounter$pregnant==1]))` unique participants with at least one care episode and a total of `r sum(encounter$pregnant==1, na.rm = T)` encounters during a pregnancy year.
5. Those missing `Modulator_trikafta_first_date` were automatically excluded from the models because they don't have a value for `Days`. This naturally means that people not eligible for Trikafta are not included in the models. 
6. 990 people with no encounters post-ETI were excluded.

```{r filtering}
# Additional quick cleaning steps
# Remove impossible lung function values
encounter$gli_fev1_ppred_rn[encounter$gli_fev1_ppred_rn < 20 |
  encounter$gli_fev1_ppred_rn > 150] <- NA
encounter$gli_fvc_ppred_rn[encounter$gli_fvc_ppred_rn < 20 |
  encounter$gli_fvc_ppred_rn > 150] <- NA
encounter$gli_fev1fvc_ppred_rn[encounter$gli_fev1fvc_ppred_rn < 20 |
  encounter$gli_fev1fvc_ppred_rn > 150] <- NA
# Filter
encounter <- encounter %>%
  filter(
    !is.na(Days),
    Days >= -365.25 * 3 & Days <= 365.25 * 2,
    hospitalized == "No", !encounterlocation %in% c("Hospital", "Home IV"),
    pregnant != 1,
    !is.na(Modulator_trikafta_first_date)
  )
# Get some summary statistics by participant
t1_participant <- encounter %>%
  group_by(eDWID) %>%
  summarise(
    `Years With Data` = length(unique(reviewyear)),
    `Number of Encounters` = n(),
    `Number of Encounters Pre-ETI` = sum(Days <= 0),
    `Number of Encounters Post-ETI` = sum(Days > 0),
    Age = mean(encounterage, na.rm = T),
    across(all_of(continuous_outcomes), \(x) mean(x, na.rm = TRUE)),
    Race = names(sort(table(Race), decreasing = TRUE))[1],
    Insurance = names(sort(table(Insurance), decreasing = TRUE))[1],
    .groups = "drop"
  )
no_post <- t1_participant$eDWID[t1_participant$`Number of Encounters Post-ETI` == 0]
encounter <- encounter %>% filter(!eDWID %in% no_post)
t1_participant <- t1_participant %>% filter(!eDWID %in% no_post)
```

# Table 1

## By person

Continuous variables were averaged across all years for each person, and categorical variables are reported as the most common value.

```{r}
t1_participant %>%
  select(-eDWID) %>%
  tbl_summary(
    by = Race,
    type = list(`Years With Data` ~ "continuous"),
    missing_text = "Missing"
  ) %>%
  add_overall()
```

# Encounter-level

```{r}
encounter %>%
  select(Race, encounterlocation, bacterialculturedone:othermicroorganisms1) %>%
  tbl_summary(by = Race, missing_text = "Missing")
```

@@

```{r functions}
# Plots
loess_plot <- function(data = encounter, outcome, facet = F) {
  p <- ggplot(data, aes(x = Days, y = !!sym(outcome), color = Race)) +
    geom_point(
      shape = ".", aes(x = Days, y = !!sym(outcome)), inherit.aes = F,
      alpha = 0.1
    ) +
    geom_smooth(method = "loess",se = F) +
    theme_classic()
  if (facet) {
    p <- p + facet_wrap(~Race)
  }
  return(p)
}
```

```{r results='asis'}
#| warning: false
invisible(lapply(continuous_outcomes, function(o) {
  cat("\n")
  cat("\n")
  cat("##", o)
  cat("\n")
  cat("\n")
  print(loess_plot(outcome = o))
}))
```

@@

# Questions

1. Do we want to combine the Asian and "Other" groups? It makes sense from a statistical point of view.
